Generalization Errors of Hard Parameter Sharing: High-dimensional Asymptotics and Implications

### 2020-10-02 arxiv/diff_cov.tex
merge examples with Section 3 and 4

>>>>

\paragraph{Example: Sample imbalance.}
The above setting assumes that the feature covariates are the same across all tasks.
Hence, the sample sizes are also the same among all tasks.
Next, we assume that the feature covariates are different and consider what happens the sample sizes are different for different tasks.
We consider a two-task case where both tasks have isotropic covariance matrices.
%	Each task has sample size $n_1 = \rho_1 \cdot p$ and $n_2 = \rho_2 \cdot p$.
	%And $X_1\in\real^{n_1\times p}, X_2\in\real^{n_2\times p}$ denote the covariates of the two tasks, respectively.
%	For the ground truth linear model, we consider a random-effects model where the two tasks differ by an independent Gaussian vector.
%	Specially, for task two, we assume that its model parameter, denoted by $\beta_2$, consists of i.i.d. Gaussian entries with mean zero and variance $\kappa^2$.
%	For task one, we assume that its model parameter, denoted by $\beta_1$, is equal to $\beta_2$ plus i.i.d. Gaussian entries with mean $0$ and variance $d^2$.
%	Therefore, the two tasks have different ground truth models. In particular, the distance between $\beta_1$ and $\beta_2$ roughly scales as $d^2 \cdot p$.
%\begin{example}[Varying model distance]
%\end{example}
Using Theorem \ref{thm_main_RMT}, we derive the prediction loss of the hard parameter sharing estimator.
Without loss of generality, we describe the result for task two and remark that a similar result for task one holds.

\begin{corollary}[Derived from Theorem \ref{thm_main_RMT}]\label{cor_MTL_loss}
	In the setting described above, assume further that a) the sample sizes $n_1$ and $n_2$ are greater than $(1 + \tau) p$, b) $\Sigma_1=\Sigma_2=\id_p$, and c) %there exists a small constant $c_0>0$ such that
	the signal and noise variances satisfy
%	\be\label{choiceofpara0}
%	p^{-1/2+c_0}\sigma^2 + p^{c_0}d^2\le \kappa^2\le p^{1-c_0} (\sigma^2 +d^2)  .  	\ee
	\be\label{choiceofpara0}
	\sigma^2 \lesssim  \kappa^2   ,\quad  d^2 \ll {\kappa^2}.
	\ee
	%Let $B, A_1, A_2$ be the global minimum of equation \eqref{eq_mtl}. Then
	The prediction loss of the hard parameter sharing estimator for task two satisfies that
	\begin{align}\label{cor_MTL_error}
	%-\left[1- \left( 1-\frac{1}{\sqrt{\rho_1}}\right)^4\right] pd^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3} +\OO(p^{-c}\sigma^2)  \le
	  \left|L(\hat{\beta}_2^{\MTL}) -\frac{\sigma^2}{\rho_1+\rho_2-1} - d^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3}\right|  & \le  \left[\left( 1+\frac{1}{\sqrt{\rho_1}}\right)^4-1\right] d^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3} \\
	& +C \left[(p^{-c_\varphi}+p^{-c_\infty/2})(\sigma^2 +d^2)+p^{-c_\infty}\kappa^2 + \frac{d^4+\sigma^2 d^2}{\kappa^2}\right],\nonumber
	 \end{align}
	 for a large constant $C>0$.
\end{corollary}

%It is well-known since the seminal work of Caruana \cite{C97} that how well multi-task learning performs depends on task relatedness. We formalize this connection in the above isotropic setting, where we can perform explicit calculations. The prediction loss of MTL has been given in Corollary \ref{cor_MTL_loss}. We can also calculate the prediction loss of STL easily using Fact \ref{lem_minv} (ii). The proof of the following claim will be given in Section \ref{app_iso_cov}.
%\begin{claim} \label{claim_STL_loss}
%Under the above isotropic setting for task two, we have that with high probability,
%\begin{align*}
%	 L(\hat{\beta}_2^{\STL}) = \frac{\sigma^2}{ \rho_2-1}  +\OO(p^{-\e}\sigma^2 )
%	 \end{align*}
%	 for any constant $\e\in (0,1/2)$.
%\end{claim}
%We now discuss two implications of Corollary \ref{cor_MTL_loss} and Claim \ref{claim_STL_loss} regarding the information transfer in MTL: the transitions from positive transfer to negative transfer with respect to varying model distance and varying sample ratio, respectively.

%\medskip
%
%\noindent{\bf Varying model distance.}
%With Corollary \ref{cor_MTL_loss} and Claim \ref{claim_STL_loss}, it is not hard to see
%One can observe that as we increase the distance $pd^2$ between $\beta_1$ and $\beta_2$, there is a transition from positive transfer to negative transfer in MTL. More precisely, the bias term $  pd^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3}$ of MTL increases as the distance between $\beta_1$ and $\beta_2$ increases, while variance reduction from STL to MTL is $\frac{\sigma^2}{ \rho_2-1}  -\frac{\sigma^2}{\rho_1+\rho_2-1}$.
%Therefore, while the variance of MTL still reduces compared to STL,
%If the bias increases more than the amount of variance reduced, we will observe negative transfer.

We apply Corollary \ref{cor_MTL_loss} to the parameter setting of Figure \ref{fig_model_shift} (the details are left to Appendix \ref{app_synthetic}). We can see that our result is able to predict positive or negative transfer  accurately and matches the empirical curve.
There are several unexplained observations near the transition threshold $0$, which are caused by the concentration error on the right-hand side of \eqref{cor_MTL_error}.
%We fix the target task and vary the source task, in particular the parameter $d$ which determines $\norm{\beta_1 - \beta_2}$.
%Figure \ref{fig_model_shift} shows the result.
%We observe that Proposition \ref{prop_dist_transition} explains most of the observations in Figure \ref{fig_model_shift}.

%The proof of Proposition \ref{prop_dist_transition} involves two parts.
%First, in equation \eqref{eq_te_var}, the positive variance reduction effect scales with $n_1 = \rho_1 p$, the number of source task data points.
%Second, we show that the negative effect of model-shift bias scales with $pd^2$, which is the expectation of $\norm{\beta_1 - \beta_2}^2$.
%The proof of Proposition \ref{prop_dist_transition} can be found in Appendix \ref{app_proof_31}.
%A key part of the analysis shows that $\hat{W}_1 / \hat{W}_2$ is roughly equal to one in the isotropic model,
%thus simplifying the general condition in Theorem \ref{thm_main_informal}.

%In classical Rademacher or VC based theory of multi-task learning, the generalization bounds are usually presented for settings where the sample sizes are equal for all tasks \cite{B00,M06,MPR16}.
%More generally, such results are still applicable when all task data are being added simultaneously.
%On the other hand, uneven sample sizes between different tasks (or even dominating tasks) have been empirically observed as a cause of negative transfer \cite{YKGLHF20}.
%For such settings, we have also observed that adding more labeled data from one task does not always help.
%Our theory accurately predicts a curious phenomenon, where increasing the sample size of the source task results in negative transfer!
%On the other hand, we have observed that adding more labeled data does not always improve performance in multi-task learning.

More precisely, in the isotropic model, we consider what happens if we vary the source task sample size. For fixed model distance $pd^2$, the bias term $  pd^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3}$ of MTL increases and converges to $pd^2$ as the sample ratio $\rho_1/\rho_2$ increases, and the variance reduction $\frac{\sigma^2}{ \rho_2-1}  -\frac{\sigma^2}{\rho_1+\rho_2-1}$ also increases and converges to $\frac{\sigma^2}{\rho_2-1}$ as $\rho_1/\rho_2$ increases. The tradeoff between these two terms depends on $\rho_1$ in a nonlinear way that is more complicated than the one in the case of varying model distance. But from our results, one can check the following phenomena. If $\frac{\sigma^2}{\rho_2-1} > pd^2$, then the transfer is always positive. On the other hand, if $\frac{\sigma^2}{\rho_2-1} < pd^2$, then there is a transition from the positive transfer to negative transfer as $\rho_1/\rho_2$ increases.
%Therefore, while the variance of MTL still reduces compared to STL,
%If the bias increases more than the amount of variance reduced, we will observe negative transfer.
Figure \ref{fig_size} provides a simulation result for such a setting (described in Appendix \ref{app_synthetic}).
%We observe that as $n_1 / n_2$ increases, there is a transition from positive to negative transfer.
There are several unexplained observations near $y = 0$ caused by the concentration error on the right-hand side of \eqref{cor_MTL_error}.

%\begin{proposition}[Source/target sample ratio]\label{prop_data_size}
%	In the isotropic model, suppose that $\rho_1 > 40$ and $\rho_2 > 110$ are fixed constants, and $\Psi(\beta_1, \beta_2) > 2/(\rho_2 - 1)$.
%	Then we have that
%	\begin{itemize}
%		\item \textbf{Positive transfer:} If $\frac{n_1}{n_2} = \frac{\rho_1}{\rho_2} < \frac{1}{\nu} \cdot \frac{1 - 2\rho_2^{-1}}{\Psi(\beta_1, \beta_2) (\rho_2 - 1) - \nu^{-1}}$, then w.h.p. $$\te(\hat{\beta}_2^{\MTL}) < \te(\hat{\beta}_2^{\STL}).$$
%		\item \textbf{Negative transfer:} If $\frac{n_1}{n_2} = \frac{\rho_1}{\rho_2} > {\nu} \cdot \frac{1 - 2\rho_2^{-1}}{\Psi(\beta_1, \beta_2) (\rho_2 - 3/2) - \nu}$, then w.h.p. $$\te(\hat{\beta}_2^{\MTL}) > \te(\hat{\beta}_2^{\STL}).$$
%	\end{itemize}
%\end{proposition}
%Proposition \ref{prop_data_size} describes the bias-variance tradeoff in terms of the sample ratio $\rho_1 / \rho_2$.

%We apply the result to the setting of Figure \ref{fig_size} (described in Appendix \ref{app_synthetic}).
%There are several unexplained observations near $y = 0$ caused by $\nu$.
%The proof of Proposition \ref{prop_data_size} can be found in Appendix \ref{app_proof_32}.

\paragraph{Example: Covariate shift.} \label{sec_covshift}
%So far we have considered the isotropic model where $\Sigma_1 = \Sigma_2$.
%This setting is relevant for settings where different tasks share the same input features such as multi-class image classification.
%In general, the covariance matrices of the two tasks may be different such as in text classification.

In our final example, we consider what happens when the covariance matrices are also different between different tasks, which are often called ``covariate shift''.
Our results apply to arbitrary ``covariate shift'' and we consider a simple setting of two tasks.
As we are going to show later, covariate shift is accurately captured by the spectrum of $\Sigma^{1/2}\Sigma^{-1/2}$.
To illustrate our result, we assume that the first half of the singular values of $\Sigma^{1/2}\Sigma^{-1/2}$ are all equal to a parameter $\lambda$ and the rest are all equal to one.
Our results imply that the prediction loss of the hard parameter sharing estimator is given as follows.
%We compare two cases: (i) when $M = \id_{p\times p}$; (ii) when $M$ has $p/2$ singular values that are equal to $\lambda$ and $p/2$ singular values that are equal to $1 / \lambda$.
%Hence, $\lambda$ measures the severity of the covariate shift.
%	Figure \ref{fig_covariate} shows a simulation of this setting by varying $\lambda$.
%	We observe that as source/target sample ratio increases, the performance gap between the two cases increases.
%	\end{example}

%By applying Lemma \ref{lem_cov_shift_informal}, we find that when $n_1 / n_2$ is large, having no covariate shift is the optimal choice provided that the determinant of $M^{\top}M$ is bounded.
%We compare different choices of $M$ that belong to the following bounded set.
%Let $\lambda_i$ be the $i$-th singular value of $M$.
%Let $\mu_{\min} < \mu < \mu_{\max}$ be fixed values that do not grow with $p$.
%{\begin{align*}
%		\cS_{\mu}\define\bigset{M \left| \prod_{i=1}^p \lambda_i \le \mu^p, \mu_{\min} \le \lambda_i\le \mu_{\max}, \text{ for all } 1\le i\le p\right.}.
%\end{align*}}
%	We assume that $\beta_1$ and $\beta_2$ are generated following the isotropic model with $d = 0$.
\begin{corollary}[Derived from Theorem \ref{thm_main_RMT}]\label{prop_covariate}
	In the setting described above, assume further that a random-effects model holds for both tasks.
	The prediction loss of $\hat{\beta}_2^{\MTL}$ for task two satisfies that
	\begin{align*}
		L(\hat{\beta}_2^{\MTL})
	\end{align*}
%	Let $g(M)$ denote the prediction loss of $\hat{\beta}_t^{\MTL}$ when $M = \Sigma_1^{1/2}\Sigma_2^{-1/2} \in\cS_{\mu}$.
%	We have that
%		\[ g(\mu\id) \le \bigbrace{1+ \bigo{{\rho_2}/{\rho_1}  }} \min_{M\in\cS_{\mu}} g(M). \]
\end{corollary}

\iffalse

 \begin{claim}\label{claim_covar_shift}
		In the setting of Proposition \ref{prop_covariate}, for any $M\in \cal S_\mu$ we have that
		\[ g(M)=(1+\OO(p^{-\e}))\cdot \sigma^2  \bigtr{\Sigma_2(X_1^{\top}X_1  + X_2^{\top}X_2)^{-1} }  \quad \text{w.h.p.} \]
	\end{claim}

%Finally, in this subsection we prove Proposition \ref{prop_covariate}, which shows that $\te(\hat{\beta}_2^{\MTL})$ is minimized approximately when $M$ is a scalar matrix, provided that there is enough source data.

\begin{proof}[Proof of Proposition \ref{prop_covariate}]

 Denote the minimizer of $g$ by
$$M_0:=\argmin_{M\in \cal S_{\mu}}g(M).$$
We now calculate $g(M_0)$.

Now using Lemma \ref{lem_cov_shift}, we obtain that with high probability,
\begin{align}\label{gvar_extra}
g(M_0)= \frac{\sigma^2}{\rho_1+\rho_2}\cdot \frac1p\tr\left( \frac{1}{a_1(M_0)\cdot M_0^\top M_0 + a_2(M_0)}\right) \cdot \left(1 +\OO(p^{-\e})\right).
\end{align}
From equation \eqref{eq_a12extra}, it is easy to obtain the following estimates on $ a_1(M)$ and $a_2(M)$ for any $M\in \cal S_\mu$:
\be\label{est_a12extra}
\frac{\rho_1-1}{\rho_1+\rho_2} < a_1(M)<  \frac{\rho_1+\rho_2-1}{\rho_1+\rho_2},\quad a_2(M) < \frac{\rho_2}{\rho_1+\rho_2}.
\ee
Inserting \eqref{est_a12extra} into \eqref{gvar_extra} and using $ M_0^\top M_0\succeq \mu_{\min}^2$, we obtain that with high probability,
\begin{align}\label{approximateteM}
\left(1+\frac{\rho_2}{(\rho_1-1)\mu_{\min}^2}\right)^{-1}h(M_0) \cdot \left(1 - \OO(p^{-\e})\right) \le g(M_0) \le h(M_0) \cdot \left(1 +\OO(p^{-\e})\right),
\end{align}
where
$$h(M_0):=\frac{\sigma^2}{(\rho_1+\rho_2)a_1(M_0)}\cdot \frac1p\tr\left( \frac{1}{M_0^\top M_0}\right) .$$
%With these two bounds, we can easily conclude \eqref{approxteM}.
%
%We have that the test error satisfies
%\be\label{approxteM}  te(M)\left(1 -  \frac{n_2}{n_1-p} \frac{1}{\lambda_p^2 + \frac{n_2}{n_1-p}}\right)  \le  \frac{\sigma^2}{n_1+n_2}\tr\left( \frac{1}{a_1M^\top M + a_2}\right) \le te(M),\ee
%where $\lambda_p$ is the smallest singular value of $p$ and
%$$te(M):= \frac{\sigma^2}{a_1(n_1+n_2)}\tr\left( \frac{1}{M^\top M}\right) .$$
%Moreover, for all $M$ satisfying \eqref{GMcons}, the minimum of $te(M)$ is attained when $M= a\id$.
By AM-GM inequality, we observe that
$$\tr\left( \frac{1}{M^\top M}\right) = \sum_{i=1}^p\frac{1}{\lambda_i^2}$$
is minimized when $\lambda_1 = \cdots\lambda_p=\mu$ under the restriction $\prod_{i=1}^p\lambda_i\le \mu^p$. Hence we get that
\be\label{AMGM} h(M_0) \le \frac{\sigma^2}{\mu^2 (\rho_1+\rho_2)a_1(M_0)}.\ee
On the other hand, when $M=\mu \id$, applying Lemma \ref{lem_cov_shift} we obtain that with high probability,
\begin{align}\label{gvar_extra2}
\begin{split}
g(\mu \id)&= \frac{\sigma^2}{\rho_1+\rho_2}\cdot \frac1p\tr\left( \frac{1}{\mu^2 a_1 (\mu\id) + a_2(\mu\id)}\right) \cdot \left(1 +\OO(p^{-\e})\right)\\
&\le \frac{\sigma^2}{\mu^2(\rho_1+\rho_2)a_1 (\mu\id)}.
\end{split}
\end{align}
Combining \eqref{est_a12extra}, \eqref{approximateteM}, \eqref{AMGM} and \eqref{gvar_extra2}, we conclude the proof.
%, we conclude that the sum $\sum_{i=1}^p\lambda_i^{-1}$ is smallest when $\lambda_1=\cdots=\lambda_p = a$.
\end{proof}
\fi

Proposition \ref{prop_covariate} shows that when the sample ratio is large, having no covariate shift gives the optimal performance for multi-task learning.
The proof of Proposition \ref{prop_covariate} can be found in Appendix \ref{app_proof_33}.
%Proposition \ref{prop_covariate} implies that when $\rho_1\gg \rho_2$, having no covariate shift is the optimal choice for choosing the source task.
%This provides evidence that covariate shift is unfavorable when there are many source task datapoints,

%\todo{} To complement the result, we show an example when the statement is not true if $n_1 \le n_2$.

%We ask: is it better to have $M$ as being close to identity, or should $M$ involve varying levels of singular values?
%Understanding this question has implications for applying normalization methods in multi-task learning \cite{LV19,CBLR18,YKGLHF20}.
%We show that if $n_1$ is much larger than $n_2$, then the optimal $M$ matrix should be proportional to identity, under certain assumptions on its range of singular values (to be formulated in Proposition \ref{prop_covariate}).
%On the other hand, if $n_1$ is comparable or even smaller than $n_2$, we show an example where having ``complementary'' covariance matrices is better performing than having the same covariance matrices.

### 2020-10-02 arxiv/experiments.tex
merge examples with Section 3 and 4

\subsection{Sample Imbalance}\label{sec_exp_sample}

@@

\subsection{Covariate Shift}

@@

\subsection{Sample Efficiency}

>>>>

\subsection{The Same Covariates Setting: Sample Efficiency}

\begin{figure}[!t]
	\begin{subfigure}[b]{0.32\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{figures/datapoints_phase_transition.pdf}
		\caption{Sample imbalance}
		\label{fig_size}
	\end{subfigure}\hfill
	\begin{subfigure}[b]{0.32\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{figures/complementary.pdf}
		\caption{Covariate shift}
		\label{fig_covariate}
	\end{subfigure}\hfill
	\begin{subfigure}[b]{0.32\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{figures/model_shift_phase_transition.pdf}
		\caption{Sample efficiency}
		\label{fig_model_shift}
	\end{subfigure}
	\caption{%Three takeaways of our theory in Section \ref{sec_insight}.
	We observe a transition from positive to negative transfer as (a) \textit{task model distance} increases and (b) source/target \textit{sample ratio} increases.
	For the special case of having the same task model, we observe in (c) that as source/target \textit{sample ratio} increases, having \textit{covariate shift} worsens the performance of MTL.
	The $y$-axis measures the loss of STL minus MTL.}
	\label{fig_model_shift_phasetrans}
\end{figure}

\subsection{The Different Covariates Setting: Sample Imbalance and Covariate Shift}\label{sec_exp_sample}

@@

@@
